---
title: "Habitat-species associations"
subtitle: "Given a number of species and habitat types in a plot, is each species significantly aggregated within habitats?"
author: "Mauro Lepore"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Habitat-species associations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
options(digits = 3)

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = FALSE,
  out.width = "70%",
  fig.align = "center",
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold"
)
```

This article shows how to determine habitat-species associations with the function [`tt_test_one()`](https://forestgeo.github.io/fgeo.habitat/reference/tt_test_one.html), developed by Sabrina Russo, Daniel Zuleta, Matteo Detto, and Kyle Harms. 

To suggest changes to the source code please file a pull request on the GitHub repository of the relevant package. You can search all functions and packages [here](https://forestgeo.github.io/fgeo/articles/fgeo.html#functions). Each function's help file links to its source code.

## Setup

```R
# install.packages("remotes")
remotes::install_github("forestgeo/fgeo.habitat")
```

For details on how to install packages from GitHub, see [this article](https://fgeo.netlify.com/2018/02/05/2018-02-05-installing-packages-from-github/).

```{r}
library(fgeo.habitat)
# For data wranging and visualization
library(tidyverse)
```

```{r}
# For reproducible results
set.seed(1014)
```

## Load census and habitat data

Here we'll use example datasets that come with __fgeo.habitat__.

```{r}
census <- luquillo_top3_sp

str(census)
```

```{r}
habitat_data <- luquillo_habitat

# Must have
names(habitat_data)

str(habitat_data)
```


To load your own data, you may run something like this:

```R
load("PATH/CENSUS_DATA.rdata")
census_data <- CENSUS_DATA

load("PATH/HABITAT_DATA.rdata")
habitat_data <- HABITAT_DATA
```

We can overview the relationship between species an habitats with a plot.

```{r}
# Tweaks
offset <- 20 / 2
habitat_centered <- mutate(
  habitat_data, 
  # Center species and habitat data
  x = x + offset, 
  y = y + offset,
  # From continuous to categorical
  habitats = as.factor(habitats)
)
```

```{r, out.width="90%", fig.widh=(6 * 0.9 / 0.7)}
ggplot(census, aes(x = gx, y = gy)) +
  geom_raster(data = habitat_centered, aes(x, y, fill = habitats)) + 
  geom_point() +
  coord_fixed() +
  facet_wrap(~sp) +
  labs(fill = "Habitat")
```

## Pick data

We'll pick alive trees, of 10 mm or more, and of sufficiently abundant species.

```{r}
pick <- filter(
  census,
  # Keep only alive
  status == "A", 
  # Keep dbh of 10 mm or more (drops missing dbh)
  dbh >= 10
)
# Count number of rows per species
pick <- add_count(pick, sp)
# Keep sufficiently abundant trees
pick <- filter(pick, n > 50)

# View
unique(select(pick, sp, n))
```

The grid size and plot dimensions will be used in more than one function. They must be consistent with the habitat data. The structure of the habitat data suggests the grid size and plot dimensions we should use ([plot details](https://forestgeo.si.edu/sites/north-america/luquillo)).

```{r}
lapply(habitat_data, range)

# For reuse
plot_dimensions <- c(320, 500)
grid_size <- 20
```

## Abundance per quadrat

`tt_test_one()` requires abundance per quadrat in a particular format, as calculated with [`abundanceperquad()`](https://s.si.edu/2JtN94p) from the original CTFSRPackage.

```{r}
# For details see ?abundanceperquad (e.g. defaults to keep trees of dbh > 10 mm)
abundance_per_quadrat <- abundanceperquad(
  pick,
  plotdim = plot_dimensions,
  gridsize = grid_size,
)$abund

str(abundance_per_quadrat, list.len = 5)
```

Now we have all we need to run `tt_test_one()` for one and multiple species.

```{r}
one_species <- unique(pick$sp)[[1]]
one_species

out_one <- tt_test_one(
  species = one_species,
  hab.index20 = habitat_data,
  allabund20 = abundance_per_quadrat,
  plotdim = plot_dimensions,
  gridsize = grid_size
)

out_one
```

To iterate over multiple species you may write a for loop, or use a functional such as `lapply()`.

```{r}
all_species <- unique(pick$sp)

out_all <- lapply(
  all_species,
  tt_test_one,
  hab.index20 = habitat_data,
  allabund20 = abundance_per_quadrat,
  plotdim = plot_dimensions,
  gridsize = grid_size
)

# For friendlier view
t(Reduce(rbind, out_all))
```

## Wrapper

We can wrap the code to run the __t__orus __t__ranslation __test__ into a convenient function, `tt_test()`, to abstract out the boilerplate code and focus on what matters most.

```{r}
tt_test <- function(sp, ...) {
  out <- lapply(sp, tt_test_one, ...)
  t(Reduce(rbind, out))
}

tt_test(
  all_species,
  hab.index20 = habitat_data,
  allabund20 = abundance_per_quadrat,
  plotdim = plot_dimensions,
  gridsize = grid_size
)
```

## Scaling up

In this section we'll use a bigger -- more realistic -- dataset.

```{r}
# Now using the pipe (`%>%`) to avoid saving intermediate results. For details
# see http://style.tidyverse.org/pipes.html.
census_bigger <- luquillo_tree6_random %>%
  filter(status == "A", dbh >= 10) %>% 
  # Keeping little abundant species (a bad idea) only to show a bigger dataset.
  add_count(sp) %>% 
  filter(n > 15)
```

Overview.

```{r, out.width="90%", fig.widh=(6 * 0.9 / 0.7)}
ggplot(census_bigger, aes(x = gx, y = gy)) +
  geom_raster(data = habitat_centered, aes(x, y, fill = habitats)) + 
  geom_point() +
  coord_fixed() +
  facet_wrap(~sp) +
  labs(fill = "Habitat")
```

The output of `tt_test()` doesn't scale up very well. With increasing number of species the resulting matrix becomes harder to visualize.

```{r}
abundance_bigger <- abundanceperquad(
  census_bigger, plotdim = plot_dimensions, gridsize = grid_size
)$abund

many_sp <- unique(census_bigger$sp)
result_bigger <- tt_test(
  many_sp,
  hab.index20 = habitat_data,
  allabund20 = abundance_bigger,
  plotdim = plot_dimensions,
  gridsize = grid_size
)

# Ackward
result_bigger
```

A solution is to convert the resulting object to a long-format dataframe, which is easier to view.

```{r}
result_long <- result_bigger %>% 
  as.data.frame() %>% 
  add_column(metric = rownames(result_bigger)) %>% 
  gather(key = "sp", value = "value", -metric)

head(result_long)
```

```R
# This view allows filtering
View(result_long)
```

We can wrap the re-formatting code into a handy function.

```{r}
tt_gather <- function(tt_test) {
  tt_test %>% 
    as.data.frame() %>% 
    add_column(metric = rownames(tt_test)) %>% 
    gather(key = "sp", value = "value", -metric)
}

result_long <- tt_gather(result_bigger)
```

## Data science workflows

Above we used `tt_gather()` to output a dataframe. Compared to a matrix, a dataframe fits better in common data science workflows. Dataframes allow using popular tools for data wrangling and visualization (from the __dplyr__ and __ggplot2__ packages). Here are some examples of what you can do.

* Filtering interesting species.

```{r}
top_sp <- census_bigger %>% 
  count(sp) %>% 
  top_n(5) %>% 
  pull(sp)
top_sp

result_top_sp <- filter(result_long, sp %in% top_sp)
# Friendly view
spread(result_top_sp, sp, value)
```

* Producing analytic summaries.

```{r}
result_top_sp %>% 
  group_by(metric) %>% 
  summarise(mean = mean(value), max = max(value), min = min(value))
```

* Producing visual summaries.

```{r, out.width="90%", fig.widh=(6 * 0.9 / 0.7)}
ggplot(result_top_sp, aes(sp, value, color = sp)) +
  geom_point() +
  facet_wrap("metric", scales = "free_y") +
  theme(axis.text.x = element_blank())
```

