---
title: "Torus translation test"
author: "Mauro Lepore"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This article shows how to use the function `torusonesp.all()`, by Sabrina Russo, Daniel Zuleta, Matteo Detto. For more information see `?torusonesp.all()`. Although the core of this function will remain largely the same, the interface to this function is in development and you can expect it to change rapidly. 

Setup.

```R
# Install the development branch 23_ttt of the package fgeo.habitat
# install.packages("remotes")
remotes::install_github("forestgeo/fgeo.habitat@23_ttt")
```

For details on how to install packages from GitHub, see [this article](https://goo.gl/dQKEeg).

```{r}
library(fgeo.habitat)
```

Example dataset from Pasoh (you should use your own data).

```{r}
census <- pasoh::pasoh_3spp
str(census)

habitat <- pasoh::pasoh_hab_index20
head(habitat)
```

To use YOUR OWN DATA, you may run something like this:

```
load("<PATH>/<CENSUS_DATA>.rdata")
census_data <- <CENSUS_DATA>

load("<PATH>/<HABITAT_DATA>.rdata")
habitat_data <- <HABITAT_DATA>
```

Abundance per quadrat.

```{r}
abundance_per_quadrat <- abundanceperquad(
  census,
  plotdim = c(1000, 500),
  gridsize = 20,
  type = 'abund'
)$abund

abundance_per_quadrat[1:10]
dim(abundance_per_quadrat)
```

Torus translation for one species.

```{r}
result_one <- torusonesp.all(
  species = "GIROPA",
  hab.index20 = habitat,
  allabund20 = abundance_per_quadrat,
  plotdim = c(1000, 500),
  gridsize = 20
)
result_one
```

Iterate over all (or a subset of) species.
```{r}
all_species <- unique(census$sp)

result_all <- lapply(
  all_species, 
  torusonesp.all,
  hab.index20 = habitat,
  allabund20 = abundance_per_quadrat,
  plotdim = c(1000, 500),
  gridsize = 20
)

# Make the output easier to view
t(Reduce(rbind, result_all))
```

